---
title: "Homework 3"
author: "Akshaya Ravikumar"
date: "1/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMBio201/Lab3/")
```

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```


# Import data
Import `Lab3_data.txt` and re-create the lab3_df and butyrate_long data frames from Lab 3 PArt 2, if they are not currently in your global environment. 
```{r}
# space to import if needed 

lab3_df <- read_delim(file = "raw_data/Lab3_data.txt", 
                      delim = "\t", escape_double = FALSE, 
                      trim_ws = TRUE, na=c("NA"),
                      col_names = TRUE, 
                      col_types = cols(participant_id = col_character(),
                                       study_week = col_character(),
                                       semester = col_character(),
                                       supplement_consumed = col_character(),
                                       quantity_compliant = col_character(),
                                       frequency = col_character(),
                                       pH_median = col_double(),
                                       pH_mean = col_double(),
                                       bristol_median = col_double(),
                                       bristol_mean = col_double(),
                                       blood_glucose_median = col_double(),
                                       blood_glucose_mean = col_double(),
                                       acetate_median = col_double(),
                                       acetate_mean = col_double(),
                                       butyrate_median = col_double(),
                                       butyrate_mean = col_double(),
                                       propionate_median = col_double(),
                                       propionate_mean = col_double())
                      )
butyrate_long <- lab3_df %>%
  # use select to drop extra columns
  select(-starts_with("ace"), -starts_with("pro"), 
         -starts_with("pH"), -starts_with("bristol"), 
         -starts_with("blood"), -ends_with("median")) %>%
  filter(semester != "Winter2015",
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT",
         quantity_compliant == "yes") %>%
  drop_na(butyrate_mean) #drop any samples with NA in named columns
```




# Question 1
[ 4 points total ]

Repeat the process used in Lab3 Part2 conduct the paired t-test for the 1xdaily group. You can use the code already written and update variable and data frame names as needed. Remember to check the assumptions. Write your conclusions regarding after the last code block. 
```{r}
# check assumptions

# sample size check 
butyrate_long %>%
  filter(study_week == "week1" | study_week == "week3", #subset long data frame in same manner as plot
         supplement_consumed == "BRMPS", 
         frequency == "1xdaily") %>%  
  group_by(frequency, study_week) %>% #set groups to what we are comparing
  summarise(sample_size = n()) #use summarise() and n() to get counts of ppl in each group

# normality check 
wk1_1x <- butyrate_long %>%
  filter(study_week == "week1", 
         supplement_consumed == "BRMPS", 
         frequency == "1xdaily") 
shapiro.test(wk1_1x$butyrate_mean) 
ggplot(wk1_1x, aes(x = butyrate_mean)) + geom_histogram()

wk3_1x <- butyrate_long %>%
  filter(study_week == "week3", 
         supplement_consumed == "BRMPS", 
         frequency == "1xdaily") 
shapiro.test(wk3_1x$butyrate_mean) 
ggplot(wk3_1x, aes(x = butyrate_mean)) + geom_histogram()

#variance test 
var.test(x = wk1_1x$butyrate_mean, 
         y = wk3_1x$butyrate_mean, 
         alternative = "two.sided") 
```

```{r}
# run test
# join these data frames back together
x1_df <- inner_join(x = wk1_1x, y = wk3_1x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  # update column names
  rename(butyrate_mean_wk1 = butyrate_mean.x,
         butyrate_mean_wk3 = butyrate_mean.y) %>%
  select(-starts_with("study_week"))

# code to run PAIRED t-test 
t.test(x = x1_df$butyrate_mean_wk1, 
       y = x1_df$butyrate_mean_wk3, 
       var.equal = TRUE, 
       paired = TRUE, 
       alternative = "less") 
```

Was there a difference in butyrate concentration between week1 and week3 in individuals who consumped BRMPS once a day? How do these results compare with the twice a day group; does this make sense in light of the figure you generated during Lab3? 

<type conclusions here>

No, there is not sufficient evidence to suggest a difference in butyrate concentration between week1 and week3 in individuals who consumed BRMPS once a day, since the p-value is greater than 0.05 (sgnificance threshold). The twice a day group was different in that there was found to be a significant difference between butyrate concentrations in week1 and week3 individuals.  The violin plot for 1xdaily generated during Lab3 also supports the conlusion of no difference, since the plots appeared to overlap heavily with minimal visible difference.

# Question 2 
[ 5 points total ]

Use Lab3_data.txt file to determine if there was a change in acetate concentrations for participants who consumed BRMPS 2xdaily. Remember to subset the data frame for quality and relevant measurements, and check all assumptions. Create a plot of the data used, save final the data frame and plot. Answer any questions listed below.
```{r}
# import and filter data frame 
acetate_long <- lab3_df %>%
  # use select to drop extra columns
  select(-starts_with("but"), -starts_with("pro"), 
         -starts_with("pH"), -starts_with("bristol"), 
         -starts_with("blood"), -ends_with("median")) %>%
  filter(semester != "Winter2015",
         supplement_consumed == "BRMPS",
         quantity_compliant == "yes",
         study_week== "week1" | study_week == "week3",
         frequency == "2xdaily") %>%
  drop_na(acetate_mean) #drop any samples with NA in named columns
dim(acetate_long)
```

What are the dimensions of the data frame?

7 columns and 133 rows

```{r}
# acetate plot
twoxdaily_plot <- acetate_long %>%
  ggplot(aes(x = study_week, 
             y = acetate_mean)) + 
  geom_violin(aes(color = study_week)) + geom_jitter(aes(color = study_week))

twoxdaily_plot
```

Is there a visual difference in acetate concentrations?

The acetate concentrations for week1 seems to be slightly more heavily concentrated towards the lower end, whereas for week3 the data is more evenly spread. But there is very minimal visual difference

```{r}
# check assumptions

#Sample Size
acetate_long %>%
  group_by(study_week) %>% #set groups to what we are comparing
  summarise(sample_size = n())

#Normality Assumption
  ace_df_week1 <- acetate_long %>%
    filter(study_week == "week1") 
  # normality test on group 
  shapiro.test(ace_df_week1$acetate_mean) #call column of dataframe with values = vector 

  ace_df_week3 <- acetate_long %>%
    filter(study_week == "week3") 
#repeat normality test with second group 
  shapiro.test(ace_df_week3$acetate_mean)
  
 
  ggplot(ace_df_week3, aes(x=acetate_mean)) +
  geom_histogram() 
qqnorm(ace_df_week3$acetate_mean); qqline(ace_df_week3$acetate_mean) 

  ggplot(ace_df_week1, aes(x=acetate_mean)) +
  geom_histogram() 
qqnorm(ace_df_week1$acetate_mean); qqline(ace_df_week1$acetate_mean) 

  
#Equal Variances
  var.test(x = ace_df_week1$acetate_mean, 
         y = ace_df_week3$acetate_mean, 
         alternative = "two.sided") 
```

Were any of the assumptions violated? Based on this, which test will you use?

The normality check was slightly violated for xdaily, but since the sample size is sufficiently large and the QQ plot follows roughly a postive straight line, it may pass the normality check. 
```{r}
# conduct statistical test
# join these data frames back together
ace_df <- inner_join(x = ace_df_week1, y = ace_df_week3,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  # update column names
  rename(acetate_mean_wk1 = acetate_mean.x,
         acetate_mean_wk3 = acetate_mean.y) %>%
  select(-starts_with("study_week")) 

t.test(x = ace_df$acetate_mean_wk1, 
       y = ace_df$acetate_mean_wk3,
       paired = TRUE, 
       var.equal = TRUE, #we set this to false based on the result of var.test() above 
       alternative = "two.sided")
```

What was the p-value of the test? What are your conclusions? 

The p-value of the test was 0.01592, which is less than 0.05, we reject the null hypothesis and suggest a change in acetate concentrations in week3 as compared to week1 in 2xdaily individuals.

```{r}
# save final data frame
write_delim(ace_df, path = "figures/hw3_question2_data.txt", delim = "\t")
# save plot
save_plot(twoxdaily_plot, 
          nrow = 2 , ncol = 2, 
          filename = "figures/Q2_plots.pdf")
```


# Question 3

[ 5 points total ]

Does breath hydrogen (H2) increase during week3 in individuals who consumed BRMPS? Start by importing the Lab3_breath.txt file, decide how you will subset the data (hint: may need to use some functions learned last week), create a plot, check assumptions, and conduct the statistical test. Write your conclusions at the end of your code, using the plot and statistical results to support your answer. Save the final data frame and plot. 
```{r}
# code block for import and subsetting
lab3_breath <- read_delim(file = "raw_data/Lab3_breath.txt", 
                      delim = "\t", escape_double = FALSE, 
                      trim_ws = TRUE, na=c("NA"),
                      col_names = TRUE, col_types = cols(
  Participant_ID = col_character(),
  Semester = col_character(),
  Supplement_consumed = col_character(),
  Quantity_compliant = col_character(),
  Frequency = col_character(),
  Study_week = col_character(),
  Measurement_type = col_character(),
  CH4_mean = col_double(),
  CH4_median = col_double(),
  CH4_adj_mean = col_double(),
  CH4_adj_median = col_double(),
  H2_mean = col_double(),
  H2_median = col_double(),
  CO2_mean = col_double(),
  CO2_median = col_double(),
  Weekly_measurements = col_double(),
  Comments = col_character()
))

H2_breathdf <- lab3_breath %>%
  # use select to drop extra columns
  select(-starts_with("CH4"), -starts_with("CO2")) %>%
  filter(Semester != "Winter2015",
         Supplement_consumed == "BRMPS",
        Study_week== "week1" | Study_week == "week3",
         Quantity_compliant == "yes") %>%
  drop_na(H2_mean)
```

```{r}
# code block for plot
H2plot <- H2_breathdf %>%
  ggplot(aes(x = Study_week, 
             y = H2_mean)) + 
  geom_violin(aes(color = Study_week)) + geom_jitter(aes(color = Study_week))

H2plot
```

```{r}
# code block for assumptions and test

#Sample Size
H2_breathdf %>%
  group_by(Study_week) %>% #set groups to what we are comparing
  summarise(sample_size = n())

#Normality Assumption
  breathwk1 <- H2_breathdf %>%
    filter(Study_week == "week1") 
  # normality test on group 
  shapiro.test(breathwk1$H2_mean) #call column of dataframe with values = vector 

  breathwk3 <- H2_breathdf %>%
    filter(Study_week == "week3") 
#repeat normality test with second group 
  shapiro.test(breathwk3$H2_mean)
 
  ggplot(breathwk3, aes(x=H2_mean)) +
  geom_histogram() 
qqnorm(breathwk3$H2_mean); qqline(breathwk3$H2_mean) 

ggplot(breathwk1, aes(x=H2_mean)) +
  geom_histogram() 
qqnorm(breathwk1$H2_mean); qqline(breathwk1$H2_mean) 

  
#Equal Variances
  var.test(x = breathwk1$H2_mean, 
         y = breathwk3$H2_mean, 
         alternative = "two.sided")

# conduct statistical test
# join these data frames back together
final_H2 <- inner_join(x = breathwk1, y = breathwk3,
                    by = c("Participant_ID", "Frequency", 
                           "Semester", "Supplement_consumed", "Quantity_compliant")) %>%
  # update column names
  rename(H2_mean_wk1 = H2_mean.x,
         H2_mean_wk3 = H2_mean.y) %>%
  select(-starts_with("Study_week")) 

t.test(x = final_H2$H2_mean_wk1, 
       y = final_H2$H2_mean_wk3,
       paired = TRUE, 
       var.equal = FALSE, #we set this to false based on the result of var.test() above 
       alternative = "less")

```

```{r}
# code block for export and save 
# save final data frame
write_delim(final_H2, path = "figures/hw3_question3_data.txt", delim = "\t")
# save plot
save_plot(H2plot, 
          nrow = 2 , ncol = 2, 
          filename = "figures/Q3_plots.pdf")
```

< Did breath hydrogen increase? write conclusions here > 

Since the p-value of 0.001136 is less than 0.05, there is sufficient evidence to suggest that breath hydrogen increased in week3 from week1.

-----
end
